Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-16 paper/paper.tex
Added implementation section

>>>>

\section{Implementation}
% GC assumptions
By the time the program starts executing, the liveness automata are created and stored in a file. 
This is re-used every time the program executes. 
% GC points - mention reduction of gc points due to e-paths
An important difference in GC of lazy language compared to an eager language is the point of 
garbage collection. In case of eager languages only \CONS~ points and function calls are 
the GC points. In a lazy language every \LET~ statement is a potential GC point. Hence we need to store 
liveness automata corresponding to every \LET~ statement in the program. This leads to an explosion 
in the number of automata. From our observation that along an {\em E-path} all program points have 
the same liveness, we can reduce the number of automata state required by sharing the same automata 
for all program points along an {\em E-path}. In the example program, we only need to create automata 
for program points \ldots, since they are the only distinct {\em E-path}s in the program. 
% Abstract description of the GC algorithm
% Liveness for cons cells & reachability for closures
During GC, if the heap cell being processed is a cons cell, we consult the liveness automaton 
corresponding to it and copy only the live paths. In case the heap cell points to a closure, 
we safely approximate the liveness to be {\{0+1\}\*} and copy everything under the closure.
% Handling of mixture of LGC & RGC
A consequence of copying everything under a closure is the fact that we can end up trying to copy 
things which are not live and corrupt the heap. This can happen if at some $GC_k$ a cons cell is copied 
using liveness. At a later stage this cons cell becomes part of a closure and hence during $GC_{k+n}$ the 
algorithm tries to copy everything under the cons cell thus copying heap locations which might not be valid. 
To avoid this behavior we either have to choose to do a full liveness based GC for closure or we should 
ensure that any non-live references are not carried across GC's. We take the latter approach and ensure that 
non-live references are set to null. After every GC, the live part of the buffer is scanned and any reference 
which is not live is set to null. The correctness of this follows from the fact that any reference which is 
decalred not live in a $GC_k$ will not be derefernced during later execution.
% A reference copied using LGC could become part of a closure in a later GC
% and copy wrong locations. Avoiding it by setting non-live references to null after every GC.
% Proof for the correctness. 
% Difference between eager and lazy GC
Another difference that is present in GC of lazy lanaguage is the presence of a print stack. In lazy 
languages the evaluation of closures is driven by the print function. Hence, during a GC there will be 
heap references that will be present on the stack which also have to copied/collected. One solution is to 
assume that all references present on the print stack are fully live and copy everything pointed by them 
akin to reachability. This is safe but copies heap cells which have already been printed. Since the print 
function is not part of the executing program its liveness automata are not considered. Therefore we 
simulate the liveness of the print function by using a flag in the heap cell. The flag ensures that only the 
required parts of a cons cell are copied. The behavior of the GC algorithm is as described below,
\begin{verbatim}
copy cons cell to new loc
if flag == false
  copy cdr part of the cell
  update cdr part of new loc
end if
\end{verbatim}
We provide an intuition about the correctness of this algorithm. Consider the following facts about the print stack 
during a GC,
\begin{enumerate}
\item The top of the stack always contains a reference to a closure.
\item All other references on the print stack point to atomic values or cons cells, it can never contain a
reference to a closure. The components of the cons cell can contain reference to closures.
\item If a reference on the print stack is a cons cell, either its car part or its cdr part will also be 
present on the stack.
\end{enumerate}
%Needs to be formatted and re-written
Sketch of the proof
Since all closures are copied fully and atomic values are copied, only cons cells need to be handled. 
Consider the following cases,
\begin{enumerate}
\item the car part of the cons cell is on the print stack. This means that the GC happened while 
the car part was being processed. The cdr part is yet to be processed and the reference to the cdr part 
is not on the print stack. Hence it is sufficient to copy only the cdr part of the cons cell. Any references 
necessary for processing the car part will be copied due to references on the print stack.
\item the cdr part of the cons cell is on the print stack. This means that GC happened while 
the cdr part was being processed.  The car part has already been processed and printed, hence we do not 
need to copy the car part. Additionally any references that are needed for printing the cdr part will 
also be copied due to the references on the print stack and hence nothing needs to be copied in this case.
\end{enumerate}
% Doing GC for references on the print stack
%  - using a flag to avoid copying non-live cells
% Optimizing GC for closures 
% - use a copied using rgc flag for closures to avoid multiple traversals
One of the reasosn why reachability based GC is faster than liveness based GC is due to the fact that 
in case we try to copy a cons cell which is already copied we do not need to traverse it any further while
 doing RGC. The same is not true in the case of LGC, as the set of liveness paths may not be similar in 
both the traversals. One optimization is to keep track of all the liveness automaton states that a particular 
cons cell has been visited with and ensure that we do not traverse it with the same state more than once. 
This optimization requires extra memory for each heap cell and potentially it can hold all the automata states. 
Therefore it is not a feasible solution. In case of lazy languages since we use reachability to copy closures, 
some cons cells might be copied using reachability. We can avoid repeated traversals of these cons cells without 
tracking the states. This only requires the knowledge of whether the cons cell was copied using reachability. 
Hence we use a flag to remember if the heap cell was copied using reachability at any time, if it is set we do 
not traverse any further.
